# WGCNA Visualizations 
> **Description**: This pipeline performs all visualizations for the WGCNA pipeline.  


**Project Initialization**
```{r ,warning = FALSE, message = FALSE}
#Sets the working directory and creates subfolders for organizing outputs.
mypath= "C:/Users/USER/Documents/Github/CRC_project/"
dir.create("output")
dir.create("plots")
dir.create("input")
```

```{r ,warning = FALSE, message = FALSE}
#load packages
library(gtools)
library(pROC)
library(ape)
library(ggdendro)
library(WGCNA)
library(stats)
library(flashClust)
library(plyr)
library(dplyr)
library(tidyr)
library(purrr)
library(tibble)
library(tidyverse)
library(gridExtra)
library(gplots)
library(ggplot2)
library(circlize)
library(ComplexHeatmap)
allowWGCNAThreads()         


```
**Prepare Data**
```{r ,warning = FALSE, message = FALSE}
#load data
dir= paste0(mypath, "output/")
# expression data
data= read.csv(paste0(mypath,"input/data_for_downstream.csv")) 
data = data |>  column_to_rownames(colnames(data)[1]) 
# differential expression file
de=   read.csv(paste0(mypath, "output/DE_sig.csv"))
de=de |> filter(abs(logFC) >= log2(3))
module_eigengenes= read.csv(paste0(mypath,"output/module_eigengenes.csv")) |>  column_to_rownames("X")


group_dist= gsub("_.*", "", colnames(data))
group_levels= unique(group_dist)

group_colors <- c("#fc8d62", "#66c2a5")
names(group_colors) <- group_levels

#change type to numeric
data[]= lapply(data, as.numeric)
datExpr= t(data) #samples become in row
#This function checks data for missing entries, entries with weights below a threshold, and zero-variance genes, 
goods <- goodSamplesGenes(datExpr, verbose = 3)
datExpr= datExpr[goods$goodSamples== TRUE, goods$goodGenes == TRUE ]

```
**Prepare Metadata**
```{r ,warning = FALSE, message = FALSE}
#metadata
metadata= data.frame(sample= colnames(data) ,
                     condition= group_dist,
                     cond_binary= as.numeric(binarizeCategoricalColumns(group_dist,
                                                             levelOrder= list(group_levels[2], group_levels[1]))[[1]]))

#metadata= read.csv("input/metadata_test.csv" )
metadata = metadata |>   column_to_rownames("sample")
metadata$condition= factor(metadata$condition, levels = group_levels)
design= model.matrix(~ 0+condition , metadata)
head(design)

```

**Set parameters and run main WGCNA function**
```{r ,warning = FALSE, message = FALSE}
#parameters for WGCNA
power= 4
minModuleSize = 20
metadata_binary= design

#Run WGCNA
net = blockwiseModules(datExpr, corType = "pearson", maxBlockSize = 5000, 
                       networkType = "signed", power = power, minModuleSize =minModuleSize,
                       mergeCutHeight = 0.25, 
                       numericLabels = F, saveTOMs = TRUE, 
                       pamRespectsDendro = FALSE, saveTOMFileBase = "TOM")

```

## **Module-phenotype Correlation Plot**  
```{r ,warning = FALSE, message = FALSE}

#Module trait correlation (visualize which module associated with what phenotype)
#We apply pearson correlation between metadata traits (conditions) and module eigen gene (1st principal component of module genes/metabolites) 

plotheatmap=function(datExpr, design, module_eigengenes){
traits= design |>  as.data.frame()
head(traits)
# Define numbers of genes and samples
nSamples <- nrow(datExpr)
nGenes <- ncol(datExpr)
module.trait.corr <- WGCNA::cor(module_eigengenes, traits, use = 'p')
module.trait.corr.pvals <- corPvalueStudent(module.trait.corr, nSamples)

#module_trait heatmap of WGCNA package
# correlations and their p-values
textMatrix = paste(signif(module.trait.corr, 2), "\n(", signif(module.trait.corr.pvals, 1), ")", 
                   sep = "")
dim(textMatrix) = dim(module.trait.corr)
par(mar = c(6, 6, 4, 6))
color= colorpanel(250, "#667F9C", "white", "#FE1B1B")

#color= greenWhiteRed(50)
# Display the correlation values within a heatmap plot
labeledHeatmap(Matrix = module.trait.corr, xLabels = gsub("condition", "", names(traits)), 
               yLabels = names(module_eigengenes), 
               ySymbols = names(module_eigengenes), colorLabels = FALSE, colors = color, 
               textMatrix = textMatrix, setStdMargins = T, cex.text = 0.8,
               zlim = c(-1, 1),xColorWidth = 1 * strheight("M"),
               yColorWidth = 1.5 * strwidth("M"),xColorOffset = strheight("M")/6, 
               yColorOffset = strwidth("M")/6, font.lab.x = 2, cex.legendLabel = 2,
               font.lab.y = 2, xLabelsAngle = 75,
               main = paste("Module-Condition Relationship"), plotLegend= TRUE)

}

#Display
plotheatmap(datExpr, design, module_eigengenes)
#Save
png("plots/heatmap_module_phenotype_cor.png", width = 4000, height = 4500, res= 600) 
plotheatmap(datExpr, design, module_eigengenes)
dev.off()

```

## **Samples Dendrogram Plot**
```{r ,warning = FALSE, message = FALSE}
#Hierarchical clustering of samples, detect outlier samples,
#and association of sample with certain trait
plotDendroAndphenotype= function(data, metadata){
#Build adjacency matrix for samples
A = adjacency(data, type = "distance")
# this calculates the whole network connectivity
k = as.numeric(apply(A, 2, sum)) - 1
# standardized connectivity
Z.k = scale(k)
# Designate samples as outlying if their Z.k value is below the threshold
thresholdZ.k = -5  # often -2.5
# the color vector indicates outlyingness (red)
outlierColor = ifelse(Z.k < thresholdZ.k, "red", "black")
# calculate the cluster tree using flahsClust or hclust
sampleTree = flashClust(as.dist(1 - A), method = "average")
# Convert traits to a color representation: where red indicates high
# values
traitColors = data.frame(numbers2colors(as.numeric(metadata$cond_binary), signed = TRUE))
#dimnames(traitColors)[[2]] = "Inflammation_lvl"
datColors = data.frame(outlier_Samples = outlierColor, Condition= traitColors)
colnames(datColors)[2]= "Conditions"

# Plot the sample dendrogram and the colors underneath.
WGCNA::plotDendroAndColors(sampleTree, groupLabels = names(datColors), colors = datColors, cex.rowText = 5,  main = "Sample dendrogram and Homogeneity of samples heatmap")

}

# Display
plotDendroAndphenotype(data,metadata)
# Save
png(paste0(mypath,"plots/WGCNA_dendrogram.png"), width = 8000, height = 6000, res= 600)
plotDendroAndphenotype(data, metadata)
dev.off()

```

## **Cluster Dendrogram Plot**
```{r ,warning = FALSE, message = FALSE}
plotDendro= function(){
plotDendroAndColors(net$dendrograms[[1]],  net$colors,
                    paste0("Merged\n", "Modules"),
                    dendroLabels = FALSE,
                    addGuide = TRUE,
                    hang= 0.03,
                    cex.colorLabels = 0.6,
                    guideHang = 0.05)
}
# Display
plotDendro()
# Save
png(paste0(mypath,"plots/dendrogram_merged_modules.png"), width = 2200, height = 2500, res= 600)
plotDendro()
dev.off()
#for error, Error in .plotOrderedColorSubplot(order = order, colors = colors, rowLabels = rowLabels,  : 
#Length of colors vector not compatible with number of objects in 'order'.
#set good samples and good genes

```

## **TOM plot (Heatmap of Correlated Features Across All Modules)**
```{r ,warning = FALSE, message = FALSE}
dissTOM= 1 - TOMsimilarityFromExpr(datExpr, power= power) #datExpr samples in rows
dendro= net$dendrograms[[1]]
moduleColorsAutomatic= net$colors

plotTOMheatmap = function() {
  #myheatcol = colorpanel(250,'gold',"orange",'darkred')
  myheatcol = colorpanel(250,'red',"orange",'lemonchiffon')
  # Transform dissTOM with a power to enhance visibility
  TOMplot(dissTOM, dendro, moduleColorsAutomatic,col= myheatcol, 
          main = "Module Heatmap Plot, All Features")
}
# Display
plotTOMheatmap()
# save
png(paste0(mypath,"plots/TOM_PLOT_module_heatmap_all.png"), width = 800, height = 600)
plotTOMheatmap()
dev.off()

```

## **MDS plot**
```{r ,warning = FALSE, message = FALSE}

dissTOM= 1 - TOMsimilarityFromExpr(datExpr, power= 5)
cmd1=cmdscale(as.dist(dissTOM),2)
# Display
plot(cmd1,col=moduleColorsAutomatic,main="MDS plot",
     xlab="Scaling Dimension 1",ylab="Scaling Dimension 2")
# Save
png(paste0(mypath,"plots/MDS_plot.png"),  width = 2800, height = 3300, res= 600) 
par(mfrow=c(1,1))
plot(cmd1,col=moduleColorsAutomatic,main="MDS plot",
     xlab="Scaling Dimension 1",ylab="Scaling Dimension 2")
dev.off()


```

## **TOM plot for Selected Features**
```{r ,warning = FALSE, message = FALSE}
# Get Differentially expressed features
genes= de$X
# subset data to have only selected proteins
datExpr_subset= datExpr[,colnames(datExpr) %in% genes]
dissTOM_subset= 1 - TOMsimilarityFromExpr(datExpr_subset, power= power)
dendro_subset = hclust(as.dist(dissTOM_subset), method = "average")
module.gene.assign= net$colors
moduleColors_subset= module.gene.assign[names(module.gene.assign)%in% genes]

png(paste0(mypath,"plots/module_heatmap_TOM_PLOT_selected.png"), width = 2800, height = 3300, res= 600)
#myheatcol = colorpanel(250,'gold',"orange",'darkred')
myheatcol = colorpanel(250,'red',"orange",'lemonchiffon')
# Transform dissTOM with a power to enhance visibility
TOMplot(dissTOM_subset, 
        dendro_subset, 
        moduleColors_subset,col= myheatcol,
        main = "Module Heatmap Plot")
dev.off()

```


```{r ,warning = FALSE, message = FALSE}
# Intramodular analysis: identifying genes with high Gene Significance (GS), i.e., strong association with the phenotype, 
#and high Module Membership (MM), i.e., strong correlation with the module eigengene (first principal component).
datKME = signedKME(datExpr, module_eigengenes)
GS.lvl= read.csv(paste0(mypath,"output/gene.trait.corr.csv")) |>  column_to_rownames("X")
modules_df= read.csv(paste0(mypath,"output/modules.csv"))
module.assign= read.csv(paste0(mypath,"output/gene_module_assignment.csv")) |>  deframe() |>  unlist()
dim(datKME)
dim(GS.lvl)


```

## **Module Membership Vs Gene Significance Scatter plot**
```{r ,warning = FALSE, message = FALSE}
# Define plotting function
plot_MM_vs_GS <- function() {
  colorOfColumn <- substring(names(datKME), 4)
  selectModules <- colnames(modules_df)
  
  # Set layout: 2 rows, N/2 columns
  par(mar = c(5, 4, 4, 2) + 0.1)
  par(mfrow = c(2, ceiling(length(selectModules) / 2)))
  
  for (module in selectModules) {
    if (!(module %in% module.assign)) {
      message(paste("Skipping module", module, "- not found in module assignment"))
      next
    }

    column <- match(module, colorOfColumn)
    restModule <- moduleColorsAutomatic == module

    if (sum(restModule) > 0) {
      verboseScatterplot(
        datKME[restModule, column],
        GS.lvl[restModule, 1],
        xlab = paste("Module Membership\n", module, "module"),
        ylab = "Feature Significance",
        main = paste("kME:", module, "vs Feature Sig."),
        col = module
      )
    } else {
      message(paste("No matching entries for module:", module))
    }
  }
}

#Display 
plot_MM_vs_GS()

#Save
png(paste0(mypath,"plots/mm_vs_sig.png"), width = 800, height = 700)
plot_MM_vs_GS()
dev.off()

```
```{r, echo=FALSE}
sessionInfo()
```
