# Module Eigengene Differentiation 
> **Description**: This pipeline performs post-WGCNA analysis by testing whether module eigengenes significantly differentiate between phenotypic groups.It uses adaptive statistical testing (parametric or non-parametric) based on normality and number of groups to assess eigengene–phenotype associations. And The results are corrected for multiple testing using FDR Benjamini-Hochberg test.

**Project Initialization**
```{r ,warning = FALSE, message = FALSE}
#Sets the working directory and creates subfolders for organizing outputs.
mypath= "C:/Users/USER/Documents/Github/CRC_project/"
dir.create("output")
dir.create("plots")
dir.create("input")

```

```{r ,warning = FALSE, message = FALSE}
#load packages
library(plyr)
library(dplyr)
library(tidyr)
library(purrr)
library(tibble)
library(tidyverse)
library(gridExtra)
library(gplots)
library(ggplot2)

```

```{r ,warning = FALSE, message = FALSE}
#load data
module_eigengenes= read.csv(paste0(mypath,"output/module_eigengenes.csv")) %>% column_to_rownames("X") 
data= read.csv(paste0(mypath,"input/data_for_downstream.csv")) 
data = data %>% column_to_rownames(colnames(data)[1])
dir= paste0(mypath, "output/")
group_dist= gsub("_.*", "", colnames(data))
group_levels= unique(group_dist)

```

## **Function for Module eigengene differentiation**
```{r ,warning = FALSE, message = FALSE}
# A function that apply the appropriate statistical test based on 
# normality and number of groups to assess differential expression 
# of module eigengenes (first principal components) between groups.
test_differentiation <- function(module_eigengenes, group_vector, output_path = NULL) {
 
  check_normality <- function(vector) {
    x <- as.numeric(vector)
    x <- x[!is.na(x)]
    if (length(x) <= 3) {
      return("Non-parametric")  
    }
    result <- tryCatch(shapiro.test(x), error = function(e) return(NULL))
    if (is.null(result)) return("Non-parametric")
    if (result$p.value > 0.05) "Parametric" else "Non-parametric"
  }
  
  # Prepare group information
  group_vector <- as.factor(group_vector)
  group_levels <- levels(group_vector)
  n_groups <- length(group_levels)
  mods <- colnames(module_eigengenes)
  
  results_list <- list()
  
  for (mod in mods) {
    x <- module_eigengenes[[mod]]
    
    test_type <- check_normality(x)
    
    if (n_groups == 2) {
      if (test_type == "Parametric") {
        res <- t.test(x ~ group_vector)
        test_used <- "t-test"
      } else {
        res <- wilcox.test(x ~ group_vector)
        test_used <- "Wilcoxon"
      }
      p <- res$p.value
    } else if (n_groups > 2) {
      if (test_type == "Parametric") {
        res <- aov(x ~ group_vector)
        p <- summary(res)[[1]][["Pr(>F)"]][1]
        test_used <- "ANOVA"
      } else {
        res <- kruskal.test(x ~ group_vector)
        p <- res$p.value
        test_used <- "Kruskal-Wallis"
      }
    } else {
      p <- NA
      test_used <- "Invalid group size"
    }
    
    results_list[[mod]] <- list(p.value = p, test = test_used)
  }
  
  # Convert results to a data frame
  res_df <- do.call(rbind, lapply(names(results_list), function(mod) {
    row <- results_list[[mod]]
    data.frame(modules = mod, p.value = row$p.value, test = row$test)
  }))
  
  # Adjust FDR
  res_df$fdr <- p.adjust(res_df$p.value, method = "BH")
  res_df$sig <- ifelse(res_df$fdr <= 0.05, "***", "")
  res_df$module <- gsub("ME", "", res_df$module)
  
  # Save if path is provided
  if (!is.null(output_path)) {
    write.csv(res_df,paste0(output_path ,"ME_differentiation.csv") , row.names = FALSE)
  }
  
  return(res_df)
}

res= test_differentiation(module_eigengenes, group_dist, dir)
head(res)
```

## **Horizontal Bar plot**
```{r ,warning = FALSE, message = FALSE}
p= ggplot(res) +
  geom_bar(aes(x = modules, y = -log10(fdr), fill = module), stat = "identity") +
  coord_flip() + # Flip coordinates to make the bar plot horizontal
  scale_fill_identity() + # Use actual colors specified in the data frame
  geom_text(aes(x =  modules  , y =-log10(fdr) , label = sig), 
            position = position_dodge(width = 0.9), vjust = -0.5, size=.5 ,color = "black") +
  theme_minimal() + 
  labs(y = "-log10(FDR)", x = "Modules") +
  ggtitle("Module Eigengene Differentiation") + # Label axes and provide a title
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5, size = 12),
        axis.text = element_text(size=12),
        axis.title=element_text(size=13) ) + 
  ylim(0,12) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black", lwd= 1)

print(p)
ggsave(paste0(mypath,"plots/module_differentiation.png"), p,width= 10, height = 7, dpi=900 )


```

